

Data analysis project example
=============================


```{r echo=FALSE, message=FALSE}

library(readr)
library(dplyr)
library(reshape2)
library(ggplot2)

library(knitr)
opts_knit$set(root.dir = "../")

```


Exploratory Analysis
--------------------


### Read data, quick look

```{r}

d <- read_csv("ds_project_example/data/credit.csv")
dim(d)
head(d)
d[[1]] <- NULL
dim(d)
head(d)
str(d)
colnames(d)

summary(d)

sapply(d, function(x) sum(is.na(x)))   ## NAs: MonthlyIncome, NumberOfDependents

sapply(d, function(x) sum(x<0, na.rm=TRUE))    ## always >= 0 

eps <- 1e-4
d %>% melt(id.vars = "SeriousDlqin2yrs") %>% tbl_df %>%
 ggplot() + geom_histogram(aes(x = value+eps, fill = as.factor(SeriousDlqin2yrs))) +
   facet_wrap(~ variable, scales = "free") + scale_x_log10()

```


### Target variable: SeriousDlqin2yrs

Person experienced 90 days past due delinquency or worse 

```{r}

summary(d$SeriousDlqin2yrs)

table(d$SeriousDlqin2yrs)

d %>% group_by(SeriousDlqin2yrs) %>% summarize(n = n()) %>% mutate(pc = n/sum(n)*100)

```



### Predictor: RevolvingUtilizationOfUnsecuredLines

Total balance on credit cards and personal lines of credit except real estate 
and no installment debt like car loans divided by the sum of credit limits
(percentage)

```{r}

summary(d$RevolvingUtilizationOfUnsecuredLines)

ggplot(d) + geom_histogram(aes(x = RevolvingUtilizationOfUnsecuredLines+eps)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + scale_x_log10()

d %>% filter(RevolvingUtilizationOfUnsecuredLines>1) %>% 
  group_by(SeriousDlqin2yrs) %>% summarize(n = n())

d %>% filter(RevolvingUtilizationOfUnsecuredLines>3) %>% 
  group_by(SeriousDlqin2yrs) %>% summarize(n = n())

ggplot(d) + geom_histogram(aes(x = RevolvingUtilizationOfUnsecuredLines)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(0,1.5))

d %>% group_by(SeriousDlqin2yrs) %>% 
  summarize(max = max(RevolvingUtilizationOfUnsecuredLines))

d %>% filter(RevolvingUtilizationOfUnsecuredLines>10) %>% 
  select(SeriousDlqin2yrs, RevolvingUtilizationOfUnsecuredLines)

```

**Notes/TODO:** clean up outliers

Options:

1. find the errors upstream, correct them
2. truncate for NN/SVM (can leave it for RF/GBM)



### Predictor: age

Age of borrower in years

```{r}

summary(d$age)

ggplot(d) + geom_histogram(aes(x = age)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free")

d %>% filter(age<18) %>% 
  group_by(SeriousDlqin2yrs) %>% summarize(n = n())

d %>% filter(age>90) %>% 
  group_by(SeriousDlqin2yrs) %>% summarize(n = n())

```



### Predictor: NumberOfTime30-59DaysPastDueNotWorse

Number of times borrower has been 30-59 days past due but no worse in the last 2 years.

```{r}

summary(d$`NumberOfTime30-59DaysPastDueNotWorse`)

ggplot(d) + geom_histogram(aes(x = `NumberOfTime30-59DaysPastDueNotWorse`)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free")

ggplot(d) + geom_histogram(aes(x = `NumberOfTime30-59DaysPastDueNotWorse`+eps)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + scale_x_log10()

table(d$`NumberOfTime30-59DaysPastDueNotWorse`)

d %>% group_by(SeriousDlqin2yrs, `NumberOfTime30-59DaysPastDueNotWorse`) %>% 
  summarize(n = n()) %>% as.data.frame

ggplot(d) + geom_histogram(aes(x = `NumberOfTime30-59DaysPastDueNotWorse`)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(0,8))

```

**Notes/TODO:** 96,98 is encoding something - needs clean up



### Predictor: DebtRatio

Monthly debt payments, alimony, living costs divided by monthy gross income (percentage)

```{r}

summary(d$DebtRatio)

ggplot(d) + geom_histogram(aes(x = DebtRatio)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free")

ggplot(d) + geom_histogram(aes(x = DebtRatio+eps)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + scale_x_log10()

ggplot(d) + geom_histogram(aes(x = DebtRatio)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(0,2))

ggplot(d) + geom_histogram(aes(x = DebtRatio)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(10,10000))

```

**Notes/TODO:** sometimes ratio, sometimes basis points? - needs cleanup



### Predictor: MonthlyIncome

Monthly income

```{r}

summary(d$MonthlyIncome)

ggplot(d) + geom_histogram(aes(x = MonthlyIncome + eps)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + scale_x_log10()

d %>% filter(MonthlyIncome < 100) %>% 
  group_by(SeriousDlqin2yrs, MonthlyIncome) %>% summarize(n = n()) %>% as.data.frame

d %>% filter(is.na(MonthlyIncome)) %>% 
  group_by(SeriousDlqin2yrs) %>% summarize(n = n()) %>% as.data.frame

ggplot(d) + geom_histogram(aes(x = MonthlyIncome + eps)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + scale_x_log10(limits = c(3e2,3e4))

d %>% filter(MonthlyIncome > 1e6) %>% select(SeriousDlqin2yrs, MonthlyIncome) %>% 
  as.data.frame

```

**Notes/TODO:** income is sometimes 0,1 or NA


### Predictor: NumberOfOpenCreditLinesAndLoans

Number of Open loans (installment like car loan or mortgage) and 
Lines of credit (e.g. credit cards)

```{r}

summary(d$NumberOfOpenCreditLinesAndLoans)

ggplot(d) + geom_histogram(aes(x = NumberOfOpenCreditLinesAndLoans)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") 

```



### Predictor: NumberOfTimes90DaysLate

Number of times borrower has been 90 days or more past due.

```{r}

summary(d$NumberOfTimes90DaysLate)

ggplot(d) + geom_histogram(aes(x = NumberOfTimes90DaysLate)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") 

ggplot(d) + geom_histogram(aes(x = NumberOfTimes90DaysLate+eps)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + scale_x_log10()

ggplot(d) + geom_histogram(aes(x = NumberOfTimes90DaysLate)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(0,10)) 

d %>% group_by(SeriousDlqin2yrs, `NumberOfTimes90DaysLate`) %>% 
  summarize(n = n()) %>% as.data.frame

```

**Notes/TODO:** 96/98



### Predictor: NumberRealEstateLoansOrLines

Number of mortgage and real estate loans including home equity lines of credit

```{r}

summary(d$NumberRealEstateLoansOrLines)

ggplot(d) + geom_histogram(aes(x = NumberRealEstateLoansOrLines)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") 

d %>% group_by(SeriousDlqin2yrs, NumberRealEstateLoansOrLines) %>% 
  summarize(n = n()) %>% as.data.frame

ggplot(d) + geom_histogram(aes(x = NumberRealEstateLoansOrLines)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(0,10))

```


### Predictor: NumberOfTime60-89DaysPastDueNotWorse

Number of times borrower has been 60-89 days past due but no worse in the last 2 years.

```{r}

summary(d$`NumberOfTime60-89DaysPastDueNotWorse`)

ggplot(d) + geom_histogram(aes(x = `NumberOfTime60-89DaysPastDueNotWorse`)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") 

d %>% group_by(SeriousDlqin2yrs, `NumberOfTime60-89DaysPastDueNotWorse`) %>% 
  summarize(n = n()) %>% as.data.frame

ggplot(d) + geom_histogram(aes(x = `NumberOfTime60-89DaysPastDueNotWorse`)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(0,6))

```

**Notes/TODO:** 96/98



### Predictor: NumberOfDependents

Number of dependents in family excluding themselves (spouse, children etc.)

```{r}

summary(d$NumberOfDependents)

ggplot(d) + geom_histogram(aes(x = NumberOfDependents)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") 

d %>% group_by(SeriousDlqin2yrs, NumberOfDependents) %>% 
  summarize(n = n()) %>% as.data.frame

ggplot(d) + geom_histogram(aes(x = NumberOfDependents)) +
  facet_grid(SeriousDlqin2yrs~., scales = "free") + xlim(c(0,6))

```

**Notes/TODO:** Deal with NAs



### Split train/test (could do CV etc.)

```{r}

d <- d %>% rename(NumberOfTime3059DaysPastDueNotWorse = `NumberOfTime30-59DaysPastDueNotWorse`,
                  NumberOfTime6089DaysPastDueNotWorse = `NumberOfTime60-89DaysPastDueNotWorse`)
colnames(d)

set.seed(123)
N <- nrow(d)
idx_train <- sample(1:N,N/2)
idx_valid <- sample(base::setdiff(1:N, idx_train), N/4)
idx_test <- base::setdiff(base::setdiff(1:N, idx_train),idx_valid)
d_train <- d[idx_train,]
d_valid <- d[idx_valid,]
d_test  <- d[idx_test,]

```


Data Cleaning
-------------

... (based on the **Notes** above)



Feature Engineering
-------------------

...




Modeling
--------


[H2O](http://h2o.ai/) is a powerful open source machine learning platform with R API


### Setup

```{r message=FALSE, results='hide'}

library(h2o)

h2o.init(max_mem_size = "4g", nthreads = -1)   ## starts Java server (R connects via REST)

dx_train <- as.h2o(d_train)  ## uploads data to H2O
dx_train$SeriousDlqin2yrs <- as.factor(dx_train$SeriousDlqin2yrs)
dx_valid <- as.h2o(d_valid)
dx_valid$SeriousDlqin2yrs <- as.factor(dx_valid$SeriousDlqin2yrs)
dx_test <- as.h2o(d_test)
dx_test$SeriousDlqin2yrs <- as.factor(dx_test$SeriousDlqin2yrs)

```

You can see/check/examine the data in the H2O (Web-based) 
UI (flow) http://localhost:54321 (or appropriate host)



### Random forest

```{r results='hide'}

system.time({
  md <- h2o.randomForest(x = 2:ncol(dx_train), y = 1, 
            training_frame = dx_train, 
            mtries = -1, ntrees = 500, max_depth = 20, nbins = 200)
})

```

```{r}

md

h2o.auc(md) 
h2o.auc(h2o.performance(md, dx_test))

```

You can examine model in the UI http://localhost:54321


### GBM

```{r results='hide'}

system.time({
  md <- h2o.gbm(x = 2:ncol(dx_train), y = 1, 
        training_frame = dx_train, validation_frame = dx_valid,
        max_depth = 15, ntrees = 500, learn_rate = 0.01, nbins = 200,
        stopping_rounds = 3, stopping_tolerance = 1e-3)
})

```

```{r}

md

h2o.auc(md)
h2o.auc(h2o.performance(md, dx_test))

```


### GBM with cross validation

```{r results='hide'}

system.time({
  md <- h2o.gbm(x = 2:ncol(dx_train), y = 1, 
          training_frame = dx_train, 
          max_depth = 15, ntrees = 500, learn_rate = 0.01, nbins = 200,
          nfolds = 5,
          stopping_rounds = 3, stopping_tolerance = 1e-3)
})


```

```{r}

md

h2o.auc(md)
h2o.auc(h2o.performance(md, dx_test))

```


### GBM with grid search

```{r results='hide'}

system.time({
  gmd <- h2o.grid("gbm", x = 2:ncol(dx_train), y = 1, 
            training_frame = dx_train, validation_frame = dx_valid,
            hyper_params = list(ntrees = 500,
                                max_depth = c(5,10,20),
                                learn_rate = c(0.01,0.1),
                                nbins = 200),
            stopping_rounds = 5, stopping_tolerance = 1e-3)
})

```

```{r}

gmd

do.call(rbind, lapply(gmd@model_ids, function(m_id) {
  mm <- h2o.getModel(m_id)
  hyper_params <- mm@allparameters
  data.frame(m_id = m_id, 
             auc = h2o.performance(mm, dx_test)@metrics$AUC,
             max_depth = hyper_params$max_depth,
             learn_rate = hyper_params$learn_rate )
})) %>% arrange(desc(auc)) 

```



### Neural network

```{r results='hide'}

system.time({
  md <- h2o.deeplearning(x = 2:ncol(dx_train), y = 1, 
          training_frame = dx_train, validation_frame = dx_valid,
          activation = "Rectifier", hidden = c(200,200), epochs = 100,
          stopping_rounds = 3, stopping_tolerance = 0)
})

```

```{r}

md

h2o.auc(md)
h2o.auc(h2o.performance(md, dx_test))

```


### Neural network with regularization (L1, L2, dropout)

```{r results='hide'}

system.time({
  md <- h2o.deeplearning(x = 2:ncol(dx_train), y = 1, 
          training_frame = dx_train, validation_frame = dx_valid,
          activation = "RectifierWithDropout", hidden = c(200,200), epochs = 100,
          input_dropout_ratio = 0.2, hidden_dropout_ratios = c(0.2,0.2),
          l1 = 1e-4, l2 = 1e-4,
          stopping_rounds = 3, stopping_tolerance = 0)
})

```

```{r}

md

h2o.auc(md)
h2o.auc(h2o.performance(md, dx_test))

```

